SPANG library: making SPARQL FAIR in controlled query sets

### 2020-05-23 paper.tex
v1.0

\begin{document}
%
\title{SPANG: constructing a controlled set of SPARQL queries with the FAIR principle}
%
\titlerunning{SPANG library}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
\author{Hirokazu Chiba}
%
\authorrunning{H. Chiba}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{Princeton University, Princeton NJ 08544, USA \and
%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
%\email{lncs@springer.com}\\
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? An increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets; however, reuse of SPARQL queries remains limited due to the lack of findability, accessibility and reusability of SPARQL queries that were once written. 
%Given that SPARQL is an important framework of the Semantic Web, it is important to consolidate the eco-system to reuse SPARQL queries written in various project.
%Here, we present a extended specification for storing and distributing SPARQL queries named SPANG, and provide a pilot study of SPANG library.
Here we propose a SPARQL extention to construct a reusable query library to enrich the Semantic Web platform.
As a use case, we constructed a reusable SPARQL query library, named SPANG library.
Our contribution is the improvement of findability and accessibility of SPARQL queries.
% These features facilitate easy access to RDF datasets by users. 
More specifically,
1) We present a metadata set for annotating a SPARQL query.
It enables to identify SPARQL query, and provides useful information fo the query.
2) We also present a mechanism to parameterize the SPARQL query.
It make each query reusable in different settings.
Further, the mechanism enables federated queries by combining multiple queries through Unix pipe. 
% Here, we developed spang command, as a client that helps users to generate queries. SPANG can create typical short queries according to the arguments specified by users like a \textit{one-liner} style in the Unix command line. It can also call SPARQL template libraries constructed in a local file system or published on the Web. 
This study make SPARQL queries more findable, accessible, interoperable, and reusable, i.e., pushing forward FAIR SPARQL.
This specification leads to the construction of eco-system to reuse SPARQL queries to enrich the use of the Semantic Web platform.
SPANG project resources is available at \url{https://spang.dbcls.jp}.
\keywords{SPARQL query \and RDF store \and template library.}
\end{abstract}
%
%
%
\section{Introduction}

An increasing number of datasets have been published in the Resource Description Framework (RDF).
SPARQL is the standardized query langage, and many Semantic Web users write queries in SPARQL, with a great amount of efforts.
Those queries are valuable resource.
However, due to the limitation of SPARQL specifications, the reuse of those queries are limited. How can we maximize the value of accumulated SPARQL queries?
%Although the SPARQL language provides a basis for exploiting RDF databases, writing a raw SPARQL code often becomes a burden for users. 
% Several approaches for support of SPARQL writing exist. One is to develop SPARQL editors that have useful functionalities 
% such as URI autocompletion~\cite{Rietveld}.
% Another aims at automatic generation of SPARQL queries~\cite{Yamaguchi}.
% Despite the efforts of previous studies, Semantic Web users remain bound to the task of SPARQL coding.
%Here, we developed SPANG, a novel SPARQL client
%that addresses this issue of SPARQL coding by simple query generation and reusing existing query templates.
%SPANG
%, with its unique features, 
%minimizes the burden of coding SPARQL,
%thereby enhancing integrative analysis of distributed RDF databases.

SPARQL is a fundamental technology for querying RDF data. However, is is not designed for reusability. In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources.

@@

The SPANG package includes the main {\tt spang} command, which can be used in the Unix command-line environment.
In general, the {\tt spang} command help users query RDF databases by dynamically generating SPARQL queries according to the command-line options or arguments. 
More specifically, {\tt spang} has two execution modes: 
1) simple mode; for a simple query, users must only specify some command-line options to generate the query without writing any SPARQL code, and
2) template mode; for a more complex query, users can generate the query using a predefined SPARQL template and parameters.
Basically, each {\tt spang} process submits a query to a specific database,
and the {\tt spang} process can be combined with other Unix processes through a Unix pipe. 
Notably, multiple {\tt spang} commands each with distinct target database can be combined through a Unix pipe, thereby realizing federated queries to multiple databases (see Section 2.4).

To reduce the burden of coding SPARQL, the SPANG package provides predefined configurations about nicknames for SPARQL endpoints, prefix declarations for URIs, and SPARQL libraries (see Section 3).

@@

The following section includes: overview of the SPANG project, specification of SPANG templates, implementation of the SPANG library, and use case of SPANSG library.

%%%%%%%%%%%%%%%%%%%%%%%%
\section{The FAIR principle}

@@

\subsection{Findable}
\begin{itemize}
    \item Assign global identifier
    \item Rich metadata
    \item Include the identifier of the data they describe
    \item Registered and indexed
\end{itemize}

\subsection{Accessible}
\begin{itemize}
    \item Retrievable by identifier
    \item Metadata are accessible
\end{itemize}

\subsection{Interoperable}
\begin{itemize}
    \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
    \item Use vocabularies that follow FAIR principles.
    \item Include qualified references to others
\end{itemize}

\subsection{Reusable}
\begin{itemize}
    \item Richly described with a plurality of accurate and relevant attributes
    \begin{itemize}
        \item released with a clear and accessible data usage license
        \item associated with detailed provenance
        \item meet domain-relevant community standards
    \end{itemize}
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG overview}
Here we overview the structure of this paper.

@@

\subsection{SPANG library}

According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

\subsection{SPANG command}

As an interface the the SPANG library, we constructed a command-line interface.

Figure~\ref{fig:spang_top} is the top page of SPANG, which includes links to GitHub, Documentation, and SPANG library.

\begin{figure}
\center
\includegraphics[width=1.0\textwidth]{spang_top.png}
\caption{SPANG web site}
\label{fig:spang_top}
\end{figure}

\subsection{Findability of SPARQL queries}
Most of the SPARQL queries are not findable from general users of the Semantic Web. The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes:
\begin{itemize}
    \item Title
    \item Description
    \item Topic
    \item Parameters: name, default value, class
    \item Target endpoint
\end{itemize}

\subsection{Example query}
An example query in the SPANG library is shown below.
The DisGeNet endpoint is specified.

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SPANG API}

@@

\subsection{Library API}

\subsection{Query API}

The following command returns the list of query libraries.

@@

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
[
  {
    "name": "disgenet",
    "title": "DisGeNET",
    "description": "DisGeNET",
    "endpoint": "http://rdf.disgenet.org/sparql/",
    "schema": "http://www.disgenet.org/web/DisGeNET/menu/rdf#schema",
    "uri": "http://localhost:7070/api/library/disgenet",
    "count": 14
  },
  {
    "name": "wikipathways",
    "title": "WikiPathways",
    "description": "WikiPathways",
    "endpoint": "http://sparql.wikipathways.org/",
    "schema": null,
    "uri": "http://localhost:7070/api/library/wikipathways",
    "count": 6
  },
  ...
]
\end{verbatim}
\end{scriptsize}
\caption{List of in query libraries obtained by API}
\label{fig:example-rdf}
\end{figure}

@@

"Precursor T-Cell Lymphoblastic Leukemia-Lymphoma"@en	"0.31"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/28671688>	"2017"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Precursor T-Cell Lymphoblastic Leukemia-Lymphoma"@en	"0.31"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/20622884>	"2010"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Colorectal Cancer"@en	"0.3"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/25944804>	"2016"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Colorectal Carcinoma"@en	"0.3"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/25944804>	"2016"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Colorectal Neoplasms"@en	"0.3"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/25944804>	"2016"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Malignant neoplasm of prostate"@en	"0.3"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/17199135>	"2007"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Prostatic Neoplasms"@en	"0.3"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/17199135>	"2007"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Lymphocyte Count measurement"@en	"0.1"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/27863252>	"2017"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Platelet Count measurement"@en	"0.1"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/27863252>	"2017"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Acute leukemia"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/21109922>	"2011"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Burkitt Lymphoma"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/21109922>	"2011"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Hematologic Neoplasms"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/21109922>	"2011"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Leukemia, Myelocytic, Acute"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/21109922>	"2011"^^<http://www.w3.org/2001/XMLSchema#gYear>
"leukemia"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/21109922>	"2011"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Acute lymphocytic leukemia"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/20622884>	"2010"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Lymphoid leukemia"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/20622884>	"2010"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Precursor Cell Lymphoblastic Leukemia Lymphoma"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/20622884>	"2010"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Breast Carcinoma"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/19146866>	"2009"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Malignant neoplasm of breast"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/19146866>	"2009"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Choriocarcinoma"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/12566248>	"2003"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Complete hydatidiform mole"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/12566248>	"2003"^^<http://www.w3.org/2001/XMLSchema#gYear>
"Gestational Trophoblastic Neoplasms"@en	"0.01"^^<http://www.w3.org/2001/XMLSchema#double>	<http://identifiers.org/pubmed/12566248>	"2003"^^<http://www.w3.org/2001/XMLSchema#gYear>
\end{verbatim}
\end{scriptsize}
\caption{Query results obtained by API}
\label{fig:example-rdf}
\end{figure}

@@

@@

\end{verbatim}
\end{scriptsize}
\caption{Search results obtained by API}
\label{fig:example-rdf}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation}
SPANG is implemented in JavaScript.

\subsubsection{SPANG template specification}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a synatax parameterization, we consider mustash notation.

\subsubsection{SPARQL template engine on the basis of AST}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG expression of SPARQL grammer originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js of SPARQL grammer is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON objecct.

@@

\subsubsection{SPARQL template library}
We implemented SPARQL queries for 10 endpoints.
We implemeted the API using Ruby.

\subsubsection{SPANG command line client}
Originally, SPANG was developed as a command line client using Perl language.
It is re-implemented in JavaScript using Node.js.
SPARQL templates can be executed in command line.

\subsubsection{SPANG Web site}
SPARQL queries can be executed in Web site.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Availability}
\subsection{Command-line client}
We developed a command-line client in JavaScript.
The usages is as follows.

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
Usage: spang2 [options] <SPARQL_TEMPLATE>

Options:
  -e, --endpoint <ENDPOINT>    target SPARQL endpoint (URL or its predifined name in SPANG_DIR/etc/endpoints,~/.spang/endpoints)
  -p, --param <PARAMS>         parameters to be embedded (in the form of "--param par1=val1,par2=val2,...")
  -o, --outfmt <FORMAT>        tsv, json, n-triples (nt), turtle (ttl), rdf/xml (rdfxml), n3, xml, html (default: "tsv")
  -a, --abbr                   abbreviate results using predefined prefixes
  -v, --vars                   variable names are included in output (in the case of tsv format)
  -S, --subject <SUBJECT>      shortcut to specify subject
  -P, --predicate <PREDICATE>  shortcut to specify predicate
  -O, --object <OBJECT>        shortcut to specify object
  -L, --limit <LIMIT>          LIMIT output (use alone or with -[SPOF])
  -F, --from <FROM>            shortcut to search FROM specific graph (use alone or with -[SPOLN])
  -N, --number                 shortcut to COUNT results (use alone or with -[SPO])
  -G, --graph                  shortcut to search for graph names (use alone or with -[SPO])
  -r, --prefix <PREFIX_FILES>  read prefix declarations (default: SPANG_DIR/etc/prefix,~/.spang/prefix)
  -n, --ignore                 ignore user-specific file (~/.spang/prefix) for test purpose
  -m, --method <METHOD>        GET or POST (default: "GET")
  -q, --show_query             show query and quit
  -f, --fmt                    format the query
  -i, --indent <DEPTH>         indent depth; use with --fmt (default: 2)
  -l, --list_nick_name         list up available nicknames of endpoints and quit
  -d, --debug                  debug (output query embedded in URL, or output AST with --fmt)
  -V, --version                output the version number
  -h, --help                   output usage information

\end{verbatim}
\end{scriptsize}
\caption{Example of input RDF data}
\label{fig:example-rdf}
\end{figure}

@@

\subsection{Simple querying with SPARQL shortcuts}

Specific command-line options, including \texttt{-S~\textit{SUBJECT}}, \texttt{-P~\textit{PREDICATE}}, \texttt{-O~\textit{OBJECT}}, \texttt{-L~\textit{LIMIT}}, and other modifiers, 
are interpreted as shortcuts for generating typical SPARQL queries. 
A simple example is:

\texttt{spang uniprot -S tax:511145 -a}

\noindent where the first argument is the target SPARQL endpoint and the following arguments are SPARQL shortcuts (the SPARQL endpoint can be specified in a URL or nickname for simplicity). 
\texttt{uniprot} is a nickname for the UniProt SPARQL endpoint \citep{UniProt} and \texttt{tax:} is a prefix for URIs of NCBI taxonomy IDs. 
This example command line searches the target database for statements that have taxonomy ID 511145 ({\it Escherichia coli} strain K-12 MG1655) as a subject. Using the \texttt{-a} option transforms the URIs in the
 search result into abbreviated forms using predefined prefix declarations. For example,\\
\texttt{\textless http://www.w3.org/2000/01/rdf-schema\#subClassOf\textgreater} 
is transformed into \texttt{rdfs:subClassOf
}.
The result is output to the standard output in the form of tab-separated values by default, as it is suitable for processing by line-oriented Unix programs.
In addition, combination of subject, predicate, and object is possible according to the following.

\texttt{spang uniprot -S tax:511145 -P up:otherName}

\noindent where predicate {\tt up:otherName} is specified to further confine the results to other names. 
For the full list of available command-line options, simply type the command {\tt spang}.
\vspace*{-5pt}

\subsection{Using SPARQL templates with parameters}

@@

\subsection{Combinatorial execution of multiple queries}

@@

SPARQL tempalting mechanisms have been implemented in various projects. Based on the templating mechanisms, SPARQL libraries have been implemented, such as Bioqueries. SPARQList is a framework to store SPRARQL templates to be used through API.

@@

Client for calling parameterized queries is still limited.
ASQC~\cite{asqc} command enables the execution of parameterized queries or queries with given variable bindings. However, it is still not possible to execute the quries publised online across the Web.

\section{Conclusion}
SPANG enables easy generation of typical queries, thereby reducing the burden of writing SPARQL. SPANG also provides a framework for reusing and sharing arbitrary queries across the Web. Moreover, it enables users to execute complex queries by combining existing query templates. SPANG, with these unique features, facilitates integrative exploitation of published RDF datasets and supports knowledge discovery.

>>>>

\begin{document}
%
\title{SPANG library: making SPARQL FAIR in controlled query sets}
%
\titlerunning{SPANG library: making SPARQL FAIR}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
%\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
\author{Hirokazu Chiba}
%
\authorrunning{H. Chiba}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{Princeton University, Princeton NJ 08544, USA \and
%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
%\email{lncs@springer.com}\\
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? 
SPARQL contributes to the standardized interface to the RDF, and maximizing the reusability of RDF; however, reusability of SPARQL itself remains to be addressed. SPARQL is a key component of the Semantic Web, thus the reusability of SPARQL is important.
FAIRification of the SPARQL is a plausible way to make SPARQL interoperable and reusable.
An increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets; however, reuse of SPARQL queries remains limited due to the lack of findability, accessibility and reusability of SPARQL queries that were once written. 
%Given that SPARQL is an important framework of the Semantic Web, it is important to consolidate the eco-system to reuse SPARQL queries written in various project.
%Here, we present a extended specification for storing and distributing SPARQL queries named SPANG, and provide a pilot study of SPANG library.
Here we propose a SPARQL extension to construct a reusable query library to enrich the Semantic Web platform.
As a use case, we constructed a reusable SPARQL query library, named SPANG library.
Our contribution is the improvement of findability and accessibility of SPARQL queries.
% These features facilitate easy access to RDF datasets by users. 
More specifically,
1) We present a metadata set for annotating a SPARQL query.
It enables to identify SPARQL query, and provides useful information fo the query.
2) We also present a mechanism to parameterize the SPARQL query.
It makes each query reusable in different settings.
Further, the mechanism enables federated queries by combining multiple queries through Unix pipe. 
% Here, we developed spang command, as a client that helps users to generate queries. SPANG can create typical short queries according to the arguments specified by users like a \textit{one-liner} style in the Unix command line. It can also call SPARQL template libraries constructed in a local file system or published on the Web. 
This study make SPARQL queries more findable, accessible, interoperable, and reusable, i.e., pushing forward FAIR SPARQL.
This specification leads to the construction of eco-system to reuse SPARQL queries to enrich the use of the Semantic Web platform.
SPANG project resources are available at \url{https://spang.dbcls.jp}.

In this study, we present the FAIRification process of SPARQL and demonstarate the usefulness of the result.

\keywords{SPARQL query \and RDF store \and template library.}

\end{abstract}
%
%
%
\section{Introduction}

SPARQL contributes to the standardized interface to RDF, and miximizing the RDF reusability. Reusability is the key to miximize the value of accumulated open data and software.
An increasing number of datasets have been published in the Resource Description Framework (RDF).
SPARQL is the standardized query langage, and many Semantic Web users write queries in SPARQL, with a great amount of efforts.
Those queries are valuable resource.
However, due to the limitation of SPARQL specifications, the reuse of those queries are limited. 
So, how can we maximize the value of accumulated SPARQL queries?
%Although the SPARQL language provides a basis for exploiting RDF databases, writing a raw SPARQL code often becomes a burden for users. 
% Several approaches for support of SPARQL writing exist. One is to develop SPARQL editors that have useful functionalities 
% such as URI autocompletion~\cite{Rietveld}.
% Another aims at automatic generation of SPARQL queries~\cite{Yamaguchi}.
% Despite the efforts of previous studies, Semantic Web users remain bound to the task of SPARQL coding.
%Here, we developed SPANG, a novel SPARQL client
%that addresses this issue of SPARQL coding by simple query generation and reusing existing query templates.
%SPANG
%, with its unique features, 
%minimizes the burden of coding SPARQL,
%thereby enhancing integrative analysis of distributed RDF databases.

SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources.

@@

% The SPANG package includes the main {\tt spang} command, which can be used in the Unix command-line environment.
% In general, the {\tt spang} command help users query RDF databases by dynamically generating SPARQL queries according to the command-line options or arguments. 
% More specifically, {\tt spang} has two execution modes: 
% 1) simple mode; for a simple query, users must only specify some command-line options to generate the query without writing any SPARQL code, and
% 2) template mode; for a more complex query, users can generate the query using a predefined SPARQL template and parameters.
% Basically, each {\tt spang} process submits a query to a specific database,
% and the {\tt spang} process can be combined with other Unix processes through a Unix pipe. 
% Notably, multiple {\tt spang} commands each with distinct target database can be combined through a Unix pipe, thereby realizing federated queries to multiple databases (see Section 2.4).

% To reduce the burden of coding SPARQL, the SPANG package provides predefined configurations about nicknames for SPARQL endpoints, prefix declarations for URIs, and SPARQL libraries (see Section 3).

@@

% The following section includes: overview of the SPANG project, specification of SPANG templates, implementation of the SPANG library, and use case of SPANSG library.

% \begin{figure}
% \center
% \includegraphics[width=1.0\textwidth]{spang_top.png}
% \caption{SPANG web site}
% \label{fig:spang_top}
% \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%
\section{FAIR principle}

The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.

@@

\subsection{Findability}
The first step is using data to find SPARQL queries. Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}

\subsection{Accessibility}
Once the user finds the required data, the user needs to know how can they be accessed, possibly including authentication and authorisation.
% \begin{itemize}
%     \item Retrievable by identifier
%     \item Metadata are accessible
% \end{itemize}

\subsection{Interoperability}
The data usually need to be integrated with other data. In addition, the data need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

\subsection{Reusability}
The ultimate goal of FAIR is to optimise the reuse of data. To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
% \begin{itemize}
%     \item Richly described with a plurality of accurate and relevant attributes
%     \begin{itemize}
%         \item released with a clear and accessible data usage license
%         \item associated with detailed provenance
%         \item meet domain-relevant community standards
%     \end{itemize}
% \end{itemize}

The process of making SPARQL FAIR ("FAIRification") can be described in multiple steps~\cite{fairification}.
Most of the requirements for findability and accessibility can be achieved at the metadata level. Interoperability and reuse requires more efforts at the SPARQL code level.

%%%%%%%%%%%%%%%%%%%%%%%%
\section{FAIRification of SPARQL}

In the previous section, we showed the core idea of the FAIR principle.

So, what is the relationship between FAIR principle and the Semantic Web?
The core technology constituting the Semantic Web is RDF, OWL, and SPARQL. RDF and OWL follows the FAIR principle. However, SPARQL is not FAIR.

So, what is the problem of the FAIRification status of SPARQL?
SPARQL is important component of the Semantic Web. The use of SPARQL query is essential for the Semantic Web community.

The key is the metadata.

\subsection{Findability of SPARQL queries}
Most of the SPARQL queries are not findable from general users of the Semantic Web. The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.

%%%%%%%%%%%%%%%%%%%%%%%%%

The core of the SPANG specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.

An example query in the SPANG library is shown below.
The DisGeNet endpoint~\cite{disgenet} is specified.

@@

\subsection{SPANG library API}
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

@@

The following command returns the list of libraries.

@@

The results is shown in Figure~\ref{fig:libs-api-out} as JSON format.

The following command returns the list of query in a library.

\texttt{curl https://spang-portal.dbcls.jp/api/library/disgenet}

The results is shown in Figure~\ref{fig:lib-api-out} as JSON format.

\subsection{SPANG query API}

@@

@@

% Figure~\ref{fig:spang_top} is the top page of SPANG, which includes links to GitHub, Documentation, and SPANG library.

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
[
  {
    "name": "disgenet",
    "title": "DisGeNET",
    "description": "DisGeNET",
    "endpoint": "http://rdf.disgenet.org/sparql/",
    "schema": "http://www.disgenet.org/web/DisGeNET/menu/rdf#schema",
    "uri": "http://localhost:7070/api/library/disgenet",
    "count": 14
  },
  {
    "name": "wikipathways",
    "title": "WikiPathways",
    "description": "WikiPathways",
    "endpoint": "http://sparql.wikipathways.org/",
    "schema": null,
    "uri": "http://localhost:7070/api/library/wikipathways",
    "count": 6
  },
  ...
]
\end{verbatim}
\end{scriptsize}
\caption{List of in query libraries obtained by API}
\label{fig:libs-api-out}
\end{figure}

@@

\end{verbatim}
\end{scriptsize}
\caption{Search results obtained by API}
\label{fig:lib-api-out}
\end{figure}

\section{Implementation details}
SPANG is implemented in JavaScript using the Node.js technology~\cite{nodejs}. Thus, it can be called in JavaScript applications, and also in the command-line environment.

\subsection{SPANG template specification}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a synatax parameterization, we consider mustash notation.

\subsection{SPARQL template engine on the basis of AST}
SPANG template engine is implemented, where each template can be transformed into abstract syntax tree, and then reconstructed into SPARQL. 
We used PEG~\cite{peg} expression of SPARQL grammer originally expressed in EBNF in the SPARQL specification. The original implementation of PEG.js~\cite{pegjs} of SPARQL grammer is provided by rdfstore-js project (https://github.com/antoniogarrote/rdfstore-js).
The PEG.js code was modified to fit the use in the case of SPANG library.
The PEG.js is a parser generator, thus, we could generate our own parser based on the PEG.js description. Thus, we obtained abstract syntax tree in the form of JSON objecct.

@@

\subsection{SPARQL template library}
We implemented SPARQL queries for 10 endpoints.

\subsection{SPANG APIs}
SPANG library API and query API are developed in Ruby on Rails. The docker version of the SPANG includes the Ruby on Rails implementation of the SPANG APIs.

\subsection{SPANG command line client}
As an interface the the SPANG library, we constructed a command-line interface.

Originally, SPANG was developed as a command line client using Perl language~\cite{spang}.
It is re-implemented in JavaScript using Node.js.
SPARQL templates can be executed in command line.

\subsection{SPANG Web site}
SPARQL queries can be executed in Web site.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Availability}

@@

\subsection{Using in JavaScript applications}
SPANG is now implemented in JavaScript, so the SPARQL queries can be executed in JavaScript code across the internet. To avid CORS retriction, it can access CDN services such as jsDelivr~\cite{jsdelivr} as shown in Figure~\ref{fig:jsdelivr}.

\begin{figure}[!t]
\begin{scriptsize}
\begin{verbatim}
spang.getTemplate('https://raw.githubusercontent.com/hchiba1/spang-library/master/li
brary/uniprot/count_org.rq', (template) => {
result = spang.query(template, 'https://sparql.uniprot.org', { abbr: true, format:
'tsv'}, (error, res, body) => { console.log(body); } );
})
\end{verbatim}
\end{scriptsize}
\caption{Call SARQL in a JavaScript application}
\label{fig:jsdelivr}
\end{figure}

% \subsection{SPANG command: simple querying with shortcuts}

% Specific command-line options, including \texttt{-S~\textit{SUBJECT}}, \texttt{-P~\textit{PREDICATE}}, \texttt{-O~\textit{OBJECT}}, \texttt{-L~\textit{LIMIT}}, and other modifiers, 
% are interpreted as shortcuts for generating typical SPARQL queries. 
% A simple example is:

% \noindent where the first argument is the target SPARQL endpoint and the following arguments are SPARQL shortcuts (the SPARQL endpoint can be specified in a URL or nickname for simplicity). 
% \texttt{uniprot} is a nickname for the UniProt SPARQL endpoint \citep{UniProt} and \texttt{tax:} is a prefix for URIs of NCBI taxonomy IDs. 
% This example command line searches the target database for statements that have taxonomy ID 511145 ({\it Escherichia coli} strain K-12 MG1655) as a subject. Using the \texttt{-a} option transforms the URIs in the
%  search result into abbreviated forms using predefined prefix declarations. For example,\\
% \texttt{\textless http://www.w3.org/2000/01/rdf-schema\#subClassOf\textgreater} 
% is transformed into \texttt{rdfs:subClassOf
% }.
% The result is output to the standard output in the form of tab-separated values by default, as it is suitable for processing by line-oriented Unix programs.
% In addition, combination of subject, predicate, and object is possible according to the following.

% \noindent where predicate {\tt up:otherName} is specified to further confine the results to other names. 
% For the full list of available command-line options, simply type the command {\tt spang}.
% \vspace*{-5pt}

\subsection{SPANG command: calling libraries}

@@

\subsection{SPANG command: combinatorial execution}

@@

SPARQL tempalting mechanisms have been implemented in various projects. Based on the templating mechanisms, SPARQL libraries have been implemented, such as Bioqueries. SPARQList is a framework to store SPARQL templates to be used through API.

@@

SPARQList~\cite{sparqlist} is a REST API server which executes a SPARQL query, transform the result into formatted data if defined. In SPARQList, configuration of the SPARQL queries such as parameters of the API, SPARQL endpoints is written in the Markdown format. SPARQList server can be considered as a repository of SPARQL queries. In terms of the FAIR principle, findability should be enhanced. That is, 1) queries are assigned with globally unique and persistent identifier; 2) queries are described with rich metadata; 3) metadata clearly and explicitly include the identifier of the data they describe; 4) queries are registered or indexed in a searchable resource.
